---
title: "PU5558-assessment1"
author: "Anonymous"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

This assessment will use the Knee Replacement CCG 2021 data set. 

The first section will answer which machine learning algorithm to choose for the 3 given questions. 

The second section will......for question no. 

## Section 1

* Estimate the post-operative EQ5D index for a patient?

* Predict how much pain a patient will have after the operation?

* Calculate how many patients have had previous surgery?

I believe that this question doesn't require a machine learning algorithm to answer the question. 

## Section 2

## Load necessary packages

```{r}

```

## Load chosen dataset

```{r}

```

## Dataset description


## Suitable machine learning algorithm for three questions:

1. Before the operation, can we estimate the post-operative EQ5D index for a patient?


2. Before the operation, can we predict how much pain a patient will have after the operation?


3. Before the operation, can we calculate how many patients have had previous surgery? 




## Model building to answer chosen question

1.	Data splitting

```{r}

```

2.	Selection and preprocessing of predictors

```{r}

```

3.	Model specification and training

```{r}

```

4.	Model evaluation

```{r}

```

## Limitations of machine learning model

