---
title: "PU5558-assessment1"
author: "Anonymous"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load necessary packages

```{r}
library(tidyverse)    # for general data science
library(tidymodels)   # for machine learning
library(vip)          # for variable importance plots
```

## Load chosen dataset

```{r}
knee_data <- read_csv("Knee Replacement CCG 2021 (2).csv")

glimpse(knee_data)

summary(knee_data)

```
```{r trying to work out all the values for 9 .....} 
#take out the variables from the data set mentioned in the AIS
knee_data%>%
  filter_all("9")

```

```{r}
rlang::last_error()
```
```{r}
summarise_all(knee_data)
```



## Dataset description

This assessment will use the Knee Replacement CCG 2021 data set.

## Suitable machine learning algorithm for three questions:

1. Before the operation, can we estimate the post-operative EQ5D index for a patient?
Supervised Learning - we have the variable, Regression. Output is numerical continuous for the index...rounding to 3 decimal places. If there is a linear relationship then you can use linear regression. We know it's supervised and we know the outcome is a numerical continuous so it must be regression. See flowchart which ML.

2. Before the operation, can we predict how much pain a patient will have after the operation?
Supervised Learning - do we have the variable??? Classification - Post-Op Q Pain


3. Before the operation, can we calculate how many patients have had previous surgery? 

I believe that this question doesn't require a machine learning algorithm to answer the question. The variable "Pre-Op Q Previous Surgery" could be used to calculate this answer by.....


## Model building to answer chosen question

1.	Data splitting

```{r}

```

2.	Selection and preprocessing of predictors

```{r}

```

3.	Model specification and training

```{r}

```

4.	Model evaluation

```{r}

```

## Limitations of machine learning model

Size of sample
PCA ??