---
title: "VIP TEST"
author: "Debbie Cannon"
date: "21/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "PU5558-assessment1"
author: "Anonymous"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction
The answers to the assessment questions are within the template sections, any additional work completed to help work out the answers are in the appendix.

## Load necessary packages

```{r All packages including those needed for the appendix}
library(tidyverse)    # for general data science
library(tidymodels)   # for machine learning
library(corrplot)     # for visualising correlation matrices
library(vip)          # for variable importance plots
library(randomForest) # for the random forest model engine
```

## Load chosen dataset

```{r}
knee_data <- read_csv("Knee Replacement CCG 2021 (2).csv")

glimpse(knee_data) #provides a general overview of the data


```
 

```{r selected columns and preprocessed data}
knee_data_col<-knee_data%>%
  select(`Pre-Op Q EQ5D Index`,`Pre-Op Q EQ VAS`,`Knee Replacement Pre-Op Q Pain`:`Knee Replacement Pre-Op Q Score`,`Post-Op Q EQ5D Index`) %>%
  drop_na() %>%    # remove rows with missing values
  unique() # keep unique row
 

```

```{r Preview the changes to the data}

glimpse(knee_data_col)

```



## Dataset description

This assessment will use the Knee Replacement CCG 2021 data set.
Note there are character (qualitative) and double (quantitative) variables within the data set. I chose variables that would be available before the operation. Only using the Post-Op EQ5D Index to train the model. There are 81 variables in the original data set, I selected 18 from the 81 variables. NOTE REMOVING MISSING VALUES AGE ETC

### Chosen potential predictor variables:

* Pre-Op EQ5D Index
* Age Band
* Gender
* Pre-Op Q EQ VAS
* Knee Replacement Pre-Op Q Pain...consecutive variables to... Knee Replacement Pre-Op Q Score

### Outcome Variable
* Post-Op Q EQ5D Index

The data set containing the 18 variables was then tidied to remove any rows with missing values. Only unique rows were kept. Finally, only rows in the Age Band and Gender without * were kept. 

I then changed the Age Band and Gender variables from characters to factors.

## Suitable machine learning algorithm for three questions:

1. Before the operation, can we estimate the post-operative EQ5D index for a patient?
Supervised Learning - we have the variable, Regression. Output is numerical continuous for the index...rounding to 3 decimal places. If there is a linear relationship then you can use linear regression. We know it's supervised and we know the outcome is a numerical continuous so it must be regression. See flowchart which ML. Linear regression, Random Forest

2. Before the operation, can we predict how much pain a patient will have after the operation?
Supervised Learning - do we have the variable??? Yes?? Classification - Post-Op Q Pain. Outcome is a classification discrete??
If it is classification it could be either logistic regression or Support Vector Machine (SVM). Logistic regression works with binary data so it can't be this algorithm because there are 5 classes?? explain briefly.


3. Before the operation, can we calculate how many patients have had previous surgery? 

I believe that this question doesn't require a machine learning algorithm to answer the question. The variable "Pre-Op Q Previous Surgery" could be used to calculate this answer by.....1 is a yes?? filter and then count.


## Model building to answer chosen question

1.	Data splitting

The data is split into a test set and a training set. The proportion of the split really depends on the context of the question. In this case the training set will contain 80% of the test data will contain 20% of the data.

```{r data splitting}

knee_data_split <- knee_data_col %>%
    initial_split(prop = 0.8, 
                  strata = `Post-Op Q EQ5D Index`) # 0.8 states the proportion of the training set. Strata ensures that the distribution of the variables in Post-Op Q EQ5D Index variable matches both the training and data set. The distribution is skewed to the right as shown in appendix 1.

knee_train <- training(knee_data_split)
knee_test <- testing(knee_data_split)

```

2.	Selection and preprocessing of predictors

```{r}

simple_rec <- knee_train %>%
  recipe(`Post-Op Q EQ5D Index`~`Pre-Op Q EQ5D Index`,`Pre-Op Q EQ VAS`,`Knee Replacement Pre-Op Q Pain`,`Knee Replacement Pre-Op Q Score`) %>%
  step_zv(all_predictors()) %>%
  step_corr(all_predictors())
```

3.	Model specification and training

```{r}
# Random forest model specification: rf_spec
rf_spec <- rand_forest() %>%
    set_mode("regression") %>%
    set_engine("randomForest")
```

```{r}

knee_wflow_rf <-workflow() %>%
           add_recipe(simple_rec) %>%
           add_model(rf_spec)
```




```{r train-model}

knee_wflow_fit <- knee_wflow_rf %>%
    fit(data = knee_train)

```


```{r}
knee_wflow_fit %>% 
  extract_fit_parsnip() %>% 
  vip(num_features = 5)
```
